The entire software and documentation produced during the Stage experience and the thesis drafting, including initial and discarded attempts are available in a public Git \href{https://github.com/avivace/dna-recombination}{repository} \cite{avivace_repo} on GitHub.

\section{Tools}
The experimental work of the Stage experience was done on a GNU/Linux Debian \texttt{buster/sid} workstation, making large of use of the shell and other tools:

\texttt{Git}, a distributed version control system, helped to keep track of every progress in the documents and the software produced.

\texttt{Python} \cite{Rossum:1995:PRM:869369} and its \texttt{IDLE} was used to quickly implement and experiment the algorithms and procedures. It's also the interface to the Gurobi interactive shell. \texttt{Ruby} was considered too.

\texttt{Gurobi Optimizer} \cite{gurobi} is a commercial optimization solver. It provides a Python interface to formulate linears problem and solve them within Python scripts.

The \texttt{TeX} typesetting engine (in the \texttt{LaTex} macros environment, with some some extensions like \texttt{BiBTex} and the \texttt{pdflatex} compiler) were used to produce the documentation, the thesis and the slides.

\section{List of produced software}
The list of produced software follows. A copy is available at \cite{avivace_repo}.
\begin{itemize}
	\item \texttt{gen.py} - generate reduced artificial instances and produce rearrangement maps (\ref{gen}).
	\item \texttt{rmapSchema.json} - proposed rearrangement map JSON schema (\ref{rmap}).
	\item \texttt{rmap.py} - example functions to parse and apply the rearrangement map format described in \ref{rmap}.
	\item \texttt{segment.py} - dynamic programming algorithm to compute every substring of a sequence.
	\item \texttt{ilp.py} - tentative Gurobi implementation of the proposed ILP formulation.
\end{itemize}

\section{Reduced artificial instance}
\label{gen}

Working on the entire genomes sequences would be prohibitive for such approach, and many of the existent solutions make assumptions on the nature of the genomes, as we've seen.

We take into consideration a reduced instance, considering a shorter sequence and only the main events (Scrambling, Inversion, Overlapping, Deletion).

A Python script which \textit{procedurally} generates an instance of the problem with given specific characteristics has been developed, it takes the following parameters to shape the desired instance:

\begin{itemize}
	\item MIC length
	\item MDS quantity
	\item Overlap size
	\item Inverse rate
\end{itemize}

The generated instance consists in:

\begin{itemize}
	\item A (randomised) MIC DNA sequence
	\item A rearrangement map containing positions, inversion flags and annotation for every MDS, Pointer in both MIC and MAC
	\item The resulting MAC sequence
\end{itemize}

Running \texttt{\$ python gen.py}, we get:
\begin{verbatim}
Generated instance:

MDS     Start   End
0       3       9
1       21      27
2       30      38
3       40      50
4       13      19

P       Start1  End1    Start2  End2
1       7       9       21      23
2       25      27      30      32
3       35      38      40      43
4       46      50      13      17

MIC     ---AAATAT----TGGAGG--ATCGGT---GTAGAATT--ATTTCGTGGA----------
               ^^    ^^^^    ^^  ^^   ^^   ^^^  ^^^   ^^^^
MAC     AAATATCGGTAGAATTTCGTGGAGG
            ^^  ^^   ^^^   ^^^^
\end{verbatim}
The parameters used were:

\begin{itemize}
	\item $60$ as MAC length
	\item A random value in the $2-5$ range as MDS quantity
	\item $0$ as Inverse rate
	\item $30\%$ as Overlap size
\end{itemize}

The pointer regions are marked in both MIC and MAC. IESs are masked for clarity. The complete annotation map is exposed through four produced objects: \texttt{MDS\_MAC}, \texttt{MDS\_Mic}, \texttt{Pointer\_MIC} and \texttt{Pointer\_MAC}, easily navigable to fetch any information about the simulated process, e.g., \texttt{Pointer\_MIC[1]["Start2"]} and \texttt{Pointer\_MIC[1]["End2"]} gives the position of the second occurrence of the first pointer section in the MIC. A standard JSON object, serialising this data, is produced, too.

\subsection{(Proposed) Rearrangement map format}
\label{rmap}
Here we show a simple rearrangement map format adopted in this work. A specification gives a coherent and reliable way interpret the events represented by the map.
A JSON schema of the format specification is proposed in \texttt{rmapSchema.json} and it's further documented on the example library that handles it (\texttt{rmap.py}).

The necessity to \textit{apply} them on genomes and produce simulations is described in \ref{correctness}.

The following events can be described:

\begin{itemize}
	\item Deletion. Implicit and trivially computable.
	\item Scrambling. The array index represents the final ordering in the MAC. Implicit, the scrambled order can be computed using the \texttt{start} and \texttt{end} positions.
	\item Overlapping of pointer sections. A pointer section is a sequence common to 2 MDSs. The pointer can appear inversed in one or both occurences.
	\item Inversion. An MDS can appear in the resulting genome as the Watson-Creek reverse complement version of the one in the MIC.
\end{itemize}

Here's how a rearrangement map in this format looks like, generated by \textit{gen.py}.

\begin{verbatim}
{
  "mac_length": 18,
  "mic_length": 60,
  "mds": [
    {
      "start": 27,
      "end": 32,
      "inverted": 0
    },
    {
      "start": 3,
      "end": 10,
      "inverted": 0
    },
    {
      "start": 14,
      "end": 24,
      "inverted": 0
    }
  ],
  "pointers": [
    {
      "start1": 30,
      "end1": 32,
      "start2": 3,
      "end2": 5
    },
    {
      "start1": 8,
      "end1": 10,
      "start2": 14,
      "end2": 16
    }
  ]
}
\end{verbatim}

This file can now be used by \texttt{rmap.py} which reproduces the events described by the map on a given genome.

\section{ILP}
A temptative \textit{pure} integer linear programming formulation follows.

\input{ilp_formulation}

\section{Preprocessing}
This part of the software computes the value for some of the variables, taking the instance as input.

Some of the defined variables are \textit{4-dimensional MIC length $\times$ MIC length $\times$ MAC length $\times$ MAC length} arrays. The necessity of a sparse data structure was immediatly clear: \textit{Sparray}, a Python module \cite{sparray} for sparse n-dimensional arrays using \textit{dictionaries} supporting any number of dimensions and any size was chosen to support these variables.

The Read-Write performance on these objects is satisfying: 15 milion integer values are written in random indexes in a 150M $\times$ 150M $\times$ 150M $\times$ 150M \textit{4d sparray} in less than 20 seconds. The data can then be accessed using a notation similar to the standard array one (\texttt{Sparray[Index1, Index2, Index3, Index4]}). \textit{Eq} and {cwc} are populated during this phase, with a naive iterative algorithm.

Here we encounter our first big limitation of a \textit{pure} linear programming approach. Populating Eq and cwc in our test instance (60 characters long MIC) was trivial but this task becomes so expensive it's infeasible with even genome sequences of more than 1000 characters.

The subsequences matching part must be approached with an high performing alignment tool like \textit{BLAST}.

\section{Gurobi Implementation}

Here is a simple Python example in order to illustrate the use of the Gurobi Python interface. The example builds a model, optimizes it, and outputs the optimal objective value. \\

\noindent Maximize $\qquad x + y + 2 z$ \\
Subject to $\quad x + 2 y + 3 z \leq  4$ $(c_0)$\quad and $x + y \geq 1 $ $(c_1)$\\
$\qquad x, y, z \qquad \text{binary}$ \\

On Python:

\begin{verbatim}
from gurobipy import *

try:
    m = Model("mip1")

    x = m.addVar(vtype=GRB.BINARY, name="x")
    y = m.addVar(vtype=GRB.BINARY, name="y")
    z = m.addVar(vtype=GRB.BINARY, name="z")

    m.setObjective(x + y + 2 * z, GRB.MAXIMIZE)

    m.addConstr(x + 2 * y + 3 * z <= 4, "c0")
    m.addConstr(x + y >= 1, "c1")

    m.optimize()

    for v in m.getVars():
        print(v.varName, v.x)

    print('Obj:', m.objVal)

except GurobiError:
    print('Error reported')
\end{verbatim}

Our model starts with variables definition:
\begin{verbatim}
[...]
MDS_Mic_Start = m.addVars(11, len(mic),
                              vtype=GRB.BINARY,
                              name="MDS_Mic_Start")
MDS_Mic_End = m.addVars(11, len(mic),
                            vtype=GRB.BINARY,
                            name="MDS_Mic_End")
Cov_Mac = m.addVars(11, len(mac), vtype=GRB.BINARY, name="Cov_Mac")
Cov_Mic = m.addVars(11, len(mic), vtype=GRB.BINARY, name="Cov_Mic")
[...]
\end{verbatim}

To define constraints we make large use of the \texttt{addConstrs} method:
\\
Add multiple constraints to a model using a Python generator expression. Returns a Gurobi tupledict that contains the newly created constraints, indexed by the values generated by the generator expression \cite{gurobi}.
\\\\
Here is how some constraints described in \ref{ilp_form} are translated in a Gurobi model: \\\\

\noindent $(3) \qquad \mathlarger{\sum_{j} MDS_{MICstart}(i,j)\leq 1 \qquad \forall i}$  \\\\\\
$(3b) \qquad \mathlarger{\sum_{j} MDS_{MACstart}(i,j)\leq 1 \qquad \forall i}$

\begin{verbatim}
m.addConstrs((sum(MDS_Mic_Start[i,a] for a in range(len(mic))) <= 1
                     for i in range(11)), name="3")

m.addConstrs((sum(MDS_Mac_Start[i,a] for a in range(len(mac))) <= 1
                     for i in range(11)), name="3b")
\end{verbatim}
\paragraph{}
\noindent $(10) \qquad \mathlarger{\sum_{i} Cov_{MAC}(i,j) \geq 1 \qquad \forall j }$ 
\begin{verbatim}
m.addConstrs((sum(Cov_Mac[i,j] for i in range(11)
                   for j in range(macl)) == macl), name="c10")
\end{verbatim}

The objective function: \\\\

\noindent $min \mathlarger{\sum_{i,j} MDS_{MACstart}(i,j)}$

\begin{verbatim}
m.setObjective((sum(MDS_Mic_Start[i,a] for a in range(0,len(mic))
                                       for i in range(0,11))),
                                       GRB.MAXIMIZE)
\end{verbatim}

11 is an upper limit of MDS quantity.

\section{Correctness}
\label{correctness}

\paragraph{Lemma.}
Suppose

$I$ to be an instance of the problem,

$P$ to be the ILP formulation associated to $I$,

$S$ to be an assignment to every variable of $P$ satisfying the constraints.

Then it's possible to build a solution for $I$ with cost equal to the objective function in $S$.

\paragraph{Proof.}
The idea is checking if \textit{S} produced by Gurobi is compatible with the known rearrangement map of the instance, i.e., if the solution proposed by Gurobi represents our DNA recombination events. "Applying" the computed map to the instance MIC should build an exact copy of the MAC.

The complete test workflow follows.

\begin{enumerate}
	\item Produce an \textit{artificial} instance \textit{I} of the problem with the software described in the \textit{Reduced artificial instance} section. By definition, this instance exhibits the rearrangement events we want to study (Scrambling, Overlapping, Deletion, Inversion) and it's a compliant instance of the formalised problem. A \textit{known} rearrangement map $R_1$ is produced too.
	\item Run the preprocess script on the generated instance and populate some of the variables of the ILP formulation associated to \textit{I}
	\item Run the Gurobi implementation of $P$ passing the populated variables. An assignment $S$ of every variable of the formulation will be computed.
	\item Build a rearrangement map $R_2$ using $S$.
	\item If $S$ is a solution then it $R_2$ maps the rearrangement events. It is possibile to simulate those events on the instance MIC and exactly obtain the MAC. $R_2$ is comparable with $R_1$.
\end{enumerate}

Step 1 is described in \ref{gen} while steps 2-4 are handled in the \texttt{ilp.py} script. Step 5 is covered in \texttt{rmap.py} where we provide a function that computes a final rearranged sequence given an initial sequence and a (compliant) rearrangement map \ref{rmap}.

This entire procedure can be pipelined and automatized to allow running a variety of tests.

\section{Conclusions}
Among other problems, inequalities like $1$ and $2$ described in \ref{ilp_form} could not be used in Gurobi, being too memory aggressive. They had to be replaced with constraints built using native gurobi methods, allowing more than bare algebra.

A \textit{pure} integer linear programming approach in this terms is clearly not successful, given the magnitude of the data to process and our choices.

A mixed approach should give better results: preprocessing the common substrings (BLAST) and producing possible instances consisting of compatible subsets of matching substrings speeds up the initial part and won't bloat the ILP formulation with huge matrices of data.

A scoring function must be designed to measure the quality of the possible maps: an ILP formulation could then optimise the problem.

Note that we (and other approaches like MIDAS \cite{midas}) use a greedy criteria for MDSs annotation: a solution with the largest possibile MDSs annotation is considered the best, requiring a 100\% MAC coverage.

As far as we know the process could actually behave differently: new speculations and the availability of transitional genomes, showing the process during intermediate phases, could change this view.